Formalizing the Kolmogorov extension Theorem in mathlib4

### 2023-09-08 Manuscript/proof_outline.tex
updated manuscript

\usepackage{environ}
\NewEnviron{killcontents}{}
\let\proof\killcontents
\let\endproof\endkillcontents

@@

\begin{abstract}
    xxx
\end{abstract}

@@

{\color{blue}
  \noindent
  TODO
  \begin{itemize}
  \item Make sure that the line of argument for code and mauscript are in line.
  \end{itemize}

}

\section{Introduction}
One of the main building blocks of modern probability theory are stochastic processes, which are usually defined as any collection of random variables -- $(X_t)_{t\in \iota}$ with $X_t$ taking values in some $\alpha_t$, say -- defined on some joint probability space $(\Omega, \mathcal A, \mathbf P)$. In order to study such processes, it is fundamental to talk about their distribution, i.e.\ a probability distribution on the product set $\prod_{t\in\iota} \alpha_t$. The usual approach to construct (the distribution of) a stochastic process works as follows: describe properties of the distribution of the stochastic process $P_J$ at some finite number of times $J = \{t_1,...,t_n\} \subseteq \iota$. The resulting family of probability measures $(P_J)_{J \subseteq \iota \text{ finite}}$ has to be {\em projective} in the sense that the projection of $P_J$ to $H\subseteq J$ has to be equal to $P_H$. In other words, when describing the distribution of the stochastic process at all times in $J$, and then forgetting (i.e.\ projecting) all properties for times in $J\setminus H$, results in the description of properties at times in $H$. One may then ask if this already gives a complete description of the process for all times. For uncountable $\iota$, e.g. $\iota = [0,\infty)$, one is tempted to be pessimistic at first sight since measures -- which describe the distribution of the stochastic process -- usually only deal well with a countable number of measurable events. However, it is the achievement of Kolmogorov that the finite-dimensional distributions in fact provide a unique description of the distribution of a stochastic process, as long as the underlying family of state spaces $(\alpha_t)_{t\in\iota}$ is nice enough. This means that these spaces are usually required to be complete and separable metric (in particular topological) spaces. This distrubution is a measure on the product-$\sigma$-field $\mathcal F :=\bigotimes_{t\in\iota} \mathcal B(\alpha_t)$ (assuming that $\mathcal B(\alpha_t)$ is the Borel $\sigma$-algebra on $\alpha_t$). Here, $\mathcal F$ is generated by finite projections and hence any element of $\mathcal F$ may only depend on at most countably many $t\in \iota$, making this a rather coarse $\sigma$-algebra. (In particular, note that this is not the Borel $\sigma$-algebra of the product topology.) The result responsible for this insight is usually denoted the {\em Kolmogorov extension theorem}, formulated in \cite{kolmogoroff1933grundbegriffe}. We note that a version of the extension result was proved by Daniell in the 1930s, but this paper was not acknowledged by the probabilists of that time \cite{aldrich2007but}. Due to his contribution, the theorem is often called the {\em Daniell-Kolmogorov extension}.

The goal of our contribution is a formalization of the proof of that theorem in {\tt mathlib4}, the mathematical library of {\tt Lean3} \cite{mathlib2020lean}. The Kolmogorov extension theorem is on the interface between measure theory and probability theory. Here, we rely on a decent amount of formalized mathematics in the measure-theory part of {\tt mathlib4} (outer measures, above all), while not requiring any specific previous formalization of probability theory. (In fact, most of our results are formulated in terms of finite rather than probability measures.)

@@

The main steps in our construction, which were previously missing in {\tt mathlib4}, are (more mathematical details are below):
\begin{enumerate}
\item a formalization of (the set-system of) semi-rings (see Definition~\ref{def:semi} below); 
\item adding some main lemmas to measure theory, e.g.\ continuity of a measure at the emptyset (see e.g.\ Lemma~\ref{l:halbextRing}, Proposition~\ref{P:stetigmass}, Lemma~\ref{l:stetigcompact} below);
\item a proof that single probability measures on a Polish space is inner regular with respect to compact sets (see Definition~\ref{def:innerreg} for a definition; in fact, we proved a slight generalization using pseudo-metric spaces, found in Lemma~\ref{L:relcoPol} below); 
\item the classical Carathéodory extension theorem, providing us a candidate for the measure which we want to construct (see Theorem~\ref{T:masseind}); 
\item the Kolmogorov extension theorem, as based on the previous steps.
\end{enumerate}
The Kolmogorov extension theorem has been previously formalized in Isabelle/HOL \cite{Immler2012}. However, this formalization only works on Polish spaces (rather than on spaces where every finite measure is inner regular with respect to compact sets, see below), and only in the case where all $\alpha_i$'s are identical. However, type theory as the basis of the formalization, as well as several mathematical concepts (e.g.\ projective families of measures or the continuity of a measure in the emptyset) are the same in both formalisations in these two interactive theorem provers.

\subsubsection*{Possible future work}
Let us describe some future projects extending {\tt mathlib4} which become possible by our contribution.

\noindent
{\em Instances of stochastic processes:} An obvious application of Kolmogorov's extension theorem is the construction of basic stochastic processes like Markov chains \cite[Section~11]{Kallenberg2020}, the Poisson process \cite[Section~13]{Kallenberg2020} and Brownian motion \cite[Section~14]{Kallenberg2020}. (Note that other Gaussian processes -- indexed by $[0, \infty)$ or $\mathbb R$ might as well be constructed the same way. In addition, fields like the Gaussien free field indexed by $\mathbb R^d$ \cite{sheffield2007gaussian} can also be given.) For these tasks, we would have to define the finite-dimensional distributions (Poisson and normally distributed, respectively), and apply the extension theorem. This task requires the formalization of multi-dimensional Poisson and normal distributions, which -- in textbooks -- is usually done using characteristic functions. Since these are not yet part of {\tt mathlib4}, we postpone this task to the future.

\noindent
{\em Sample-path properties:} The Kolmogorov extension gives the existence of a distribution of a stochastic process $(X_t)_{t\in\iota}$ with certain properties. Extra work is needed in order to show that -- on the same probability space -- we can as well define a version (i.e.\ another process $(Y_t)_{t\in\iota}$ with $\mathbf P(X_t = Y_t) = 1$ for all $t\in\iota$) which is right-continuous with left limits (for the Poisson process) or continuous (for Brownian motion), respectively. For the former, this follows at least from some general principles of Markov processes (e.g.\ \cite[Theorem 4.3.6]{EthierKurtz1986}). For the latter, this requires formalization of the Kolmogorov-Chentsov criterion 
\cite[Theorem~4.23]{Kallenberg2020}.

\section{Formalization of the Daniell-Kolmogorov extension}
We start by stating the exact result. We are going to formulate the main result in a modern fashion, as e.g.\ found in Theorem 2.2 of \cite{rao1971projective}, Theorem 7.7.1 of Volume~2 of \cite{bogachev2007measure}, Theorem 15.26 of \cite{guide2006infinite}, or \cite{border1998expository}. Note that these formulations split general assumptions on the underlying space(s) (e.g.\ a metric property) from the property which is needed in the proof (inner regularity with respect to compact sets). Other -- highly readable -- references such as \cite{Billingsley1995} state the extension theorem only in special cases such as $\alpha_t = \mathbb R$ for all $t$. 
%In addition, there are still new proofs found for the general extension theorem \cite{chin2019new}, but they require some results not yet formalized in mathlib4 (Borel isomorphism of $\mathbb R$ to $\{0,1\}^{\mathbb N}$), so we do not follow this route.  
Whenever we feel necessary, we rely on a standard textbooks on Probability Theory \cite{Klenke2013, Kallenberg2020} for general purpose.

Before we can state the result, we begin with some mathematical notions. We start off slowly from metric spaces, but quickly introduce projective families of measures. In all definitions, $\alpha$ will be some set (or type, since {\tt Lean4} is a dependently typed language, \leanline{α : Type _}).

\begin{definition}[Metric and topological spaces]
  \begin{enumerate}
  \item \sloppy A pseudo-metric on $\alpha$ is a symmetric map $r : \alpha
    \times \alpha \to [0,\infty)$ satisfying the triangle inequality,
      i.e.\ $r(x,z) \leq r(x,y) + r(y,z)$ for all $x,y,z\in\alpha$.
      \\ If $r$ also satisfies $r(x,y) = 0$ iff $x=y$, we call it a
      metric.  \\ If $r : \alpha \times \alpha \to [0,\infty]$
      (i.e.\ $r(x,y) = \infty$ is allowed), we call $r$ an extended
      (pseudo-)metric.
    \item Let $r$ be an (extended pseudo-)metric. A sequence $x_1,
      x_2,... \in \alpha$ is called Cauchy if for all $\varepsilon>0$
      there is $N\in\mathbb N$ such that $r(x_n, x_m) < \varepsilon$
      for all $m,n > N$. The (pseudo-extended) metric is called
      complete if every Cauchy-sequency has a limit in
      $\alpha$.
    \item Some\footnote{We denote by $2^\alpha$ the power set of
      $\alpha$, i.e.\ the set of all subsets of $\alpha$.} $\mathcal O
      \subseteq 2^\alpha$ is called a topology, if it satisfies (i)
      $\emptyset, \alpha \in \mathcal O$, (ii) $\mathcal O$ is stable
      under arbitrary unions, i.e.\ if $A_i \in \mathcal O$ for all
      $i\in\iota$ and $\iota$ is arbitrary, then $\bigcup_{i\in \iota}
      A_i \in \mathcal O$, (iii) $\mathcal O$ is stable under finite
      intersections, i.e.\ if $A_1,...,A_n \in \mathcal O$, then
      $\bigcap_{i\in \iota} A_i \in \mathcal O$.
  \end{enumerate}
\end{definition}

\begin{remark}[Metric and topological spaces in {\tt mathlib4}] \mbox{}
  \begin{enumerate}
    \item \sloppy In \leanline{Lean4}, we have the \leanline{class
      PseudoEMetricSpace (α : Type u) extends EDist}, which comes with
      the components, \leanline{edist_self}, \leanline{edist_comm},
      \leanline{edist_triangle} providing the properties of the
      pseudo-metric, and \leanline{toUniformSpace},
      \leanline{uniformity_edist} defining a uniform space from the
      extended pseudo-metric. Such a space does not come with a
      metric, but with a filter on $\alpha \times \alpha$, which
      describes which points in $\alpha$ are near. For example, the
      diagonal of $\alpha \times \alpha$ is a subset of all sets in
      the uniformity; see \cite{james2013topologies} for details. We
      note that uniform spaces with a countably generated uniformity
      filter is pseudometrizable, i.e.\ there exists a
      pseudo-metric-space structure that generates the same
      uniformity; see \leanline{UniformSpace.pseudoMetrizableSpace},
      which formalizes a result stated in
      \cite{melikhov2011metrizable}.
    \item \sloppy As for topological spaces, we have \leanline{class
      TopologicalSpace(α : Type u) : Type u}, which comes with
      \leanline{isOpen}, \leanline{isOpen_univ},
      \leanline{isOpen_inter} and \leanline{isOpen_sUnion}, which are
      exactly the properties of a topological space described above.
    \item Any (extended pseudo-)metric on $\alpha$ defines a topology,
      namely the topology generated by $\mathcal H := \{\{y:
      r(x,y)<\varepsilon\}: x \in \alpha, \varepsilon \in
      (0,\infty)\}$. Actually, in \leanline{mathlib4}, as we saw
      above, every extended pseudo-metric defines a uniform space, and
      a uniform space is a \leanline{class UniformSpace(α : Type u)
        extends TopologicalSpace , UniformSpace.Core} with xxx
      \leanline{to_topological_space} giving a topological space,
      \leanline{to_core} which stores the uniformity filter, and
      \leanline{is_open_uniformity}, which gives all open sets. In
      particular, a uniform space defines a topological space, which
      can be used by the typeclass system.
  \end{enumerate}
\end{remark}

\begin{remark}[Generated topology]
  \begin{enumerate}
  \item The intersection of any number of topologies is again a
    topology. For this reason, if $\mathcal H \subseteq 2^\alpha$, we
    define the topology $\mathcal O := \bigcap_{\mathcal F \supseteq
      \mathcal H \text{ topology}} \mathcal F$; see
    \leanline{TopologicalSpace.generateFrom}. This is called the
    topology generated by $\mathcal H$. If $\mathcal H$ is closed
    under finite intersections, we call $\mathcal H$ a basis for
    $\mathcal O$; see \leanline{TopologicalSpace.IsTopologicalBasis}.
  \item \sloppy We call the topology (generated from an extended
    pseudo-metric) separable (see
    \leanline{TopologicalSpace.SeparableSpace}) if there is a
    countable $s \subseteq \iota$ such that $\inf\{r(x,y) : y \in s\}
    = 0$ for all $x\in\alpha$. \\ If there is a countable basis of the
    topology, it is separable; see
    \leanline{TopologicalSpace.SecondCountableTopology.to_separableSpace}.
  \end{enumerate}
\end{remark}

\noindent
Finally, we can introduce measures. Still, we need even more
set-systems, i.e.\ $\sigma$-algebras.

\begin{definition}[$\sigma$-algebras and measures]
  \begin{enumerate}
  \item 
    We call $\mathcal F \subseteq 2^\alpha$ a $\sigma$-algebra (on
    $\alpha$) if (i) $\alpha \in \mathcal F$, (ii) $\mathcal F$ is
    stable under complements, i.e.\ $A \in \mathcal F \Rightarrow A^c
    \in \mathcal F$, (iii) $\mathcal F$ is stable under countable
    unions, i.e.\ $A_1, A_2,... \in \mathcal F \Rightarrow
    \bigcup_{n=1}^\infty A_n \in \mathcal F$. We call $(\alpha,
    \mathcal F)$ a measurable space.
  \item For some $\sigma$-algebra $\mathcal F$ on $\alpha$, a function
    $\mu : \mathcal F \to [0, \infty]$ is called a measure,
    if\footnote{Since $\alpha \in \mathcal F$ and $\mathcal F$ is
      stable under complements, we also have that $\emptyset \in
      \mathcal F$} (i) $\mu(\emptyset) = 0$, (ii) $\mu$ is countably
    additive, i.e.\ for $A_1, A_2,... \in \mathcal F$ pairwise
    disjoint, we have $\mu\Big( \bigcup_{n=1}^\infty A_n\Big) =
    \sum_{n=1}^\infty \mu(A_n)$. We call $(\alpha, \mathcal F, \mu)$ a
    measure space. \\ In addition, $\mu$ is called finite if
    $\mu(\alpha) < \infty$ and a probability measure if
    $\mu(\alpha)=1$.
  \end{enumerate}
\end{definition}

\begin{remark}[Measur(abl)e spaces in \leanline{mathlib4}]
  \begin{enumerate}
  \item \sloppy The \leanline{class MeasurableSpace(α : Type u_7) :
    Type u_7} is very similar to a topological space in
    \leanline{mathlib4}, since it comes with
    \leanline{MeasurableSet'}, \leanline{measurableSet_empty},
    \leanline{measurableSet_compl}, and
    \leanline{measurableSet_iUnion}.
  \item For a measure space, we require a measurable space as a
    variable, and add the set function: \leanline{structure
      MeasureTheory.Measure(α : Type u_6) [inst : MeasurableSpace α]
      extends MeasureTheory.OuterMeasure}. Here $\sigma$-additivity of
    a \leanline{Measure α} reads xxx note on outer measures
\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  m_iUnion : ∀ ⦃f : ℕ → Set α⦄,
    (∀ (i : ℕ), MeasurableSet (f i)) →
    Pairwise (Disjoint on f) →
    ↑toOuterMeasure (⋃ (i : ℕ), f i) =
    ∑' (i : ℕ), ↑toOuterMeasure (f i)
\end{minted}    
  \end{enumerate}
\end{remark}

@@

Using \leanline{variable {α : Type _} {C : Set (Set α)} {s t : Set α}
}, we define a semi-ring as follows:

@@

Recall that \leanline{Lean4} indicates a coercion by \leanline{↑},
which in this case is from \leanline{Finset} to \leanline{Set}.

>>>>

%\usepackage{environ}
%\NewEnviron{killcontents}{}
%\let\proof\killcontents
%\let\endproof\endkillcontents

@@

\begin{abstract}
  We present a formalization of the Kolmogorov extension theorem,
  which is a main building block in construction the (distribution of)
  stochastic processes with arbitrary index sets in Lean4. Our
  approach is based on mathlib, the mathematical library for Lean4. On
  our way, we provide a formalization of the Caratheodory extension
  theorem, which allows to extend a set function on a semi-ring to a
  proper measure on the $\sigma$-algebra generated by the semi-ring. A
  slight generalization of the classical Kolmogorov extension theorem
  allows us to construct the (distribution of) a stochastic process on
  a complete and separable pseudo-metric space (rather than a metric
  space).
\end{abstract}

@@

{\color{blue}
  \noindent
  TODO
  \begin{itemize}
  \item Make sure that the line of argument for code and mauscript are in line.
  \end{itemize}
}

\section{Introduction}
One of the main building blocks of modern probability theory are stochastic processes, which are usually defined as any collection of random variables -- $(X_t)_{t\in \iota}$ with $X_t$ taking values in some $\alpha_t$ for all $t\in\iota$ -- defined on some joint probability space $(\Omega, \mathcal A, \mathbf P)$.  (As usual, we will refer to $\iota$ as the index set of times.) In order to study such processes, it is fundamental to talk about their joint distribution, i.e.\ a probability distribution on the product set $\prod_{t\in\iota} \alpha_t$. The usual approach to construct (the distribution of) a stochastic process works as follows: describe properties of the distribution of the stochastic process $P_J$ at some arbitrary but finite number of times $J = \{t_1,...,t_n\} \subseteq \iota$. The resulting family of probability measures $(P_J)_{J \subseteq \iota \text{ finite}}$ has to be {\em projective} in the sense that the projection of $P_J$ to $H\subseteq J$ has to be equal to $P_H$. In other words, when describing the distribution of the stochastic process at all times in $J$, and then forgetting (i.e.\ projecting) all properties for times in $J\setminus H$, results in the description of properties at times in $H$. One may then ask if this already gives a complete description of the process for all times. For uncountable $\iota$, e.g. $\iota = [0,\infty)$, one is tempted to be pessimistic at first sight since measures -- which describe the distribution of the stochastic process -- usually only deal well with a countable number of measurable events. However, it is the achievement of Kolmogorov that the finite-dimensional distributions in fact provide a unique description of the distribution of a stochastic process, as long as the underlying family of state spaces $(\alpha_t)_{t\in\iota}$ is nice enough (Polish, i.e.\ a separable topological space which can be metrized by a complete metric, for example). Assuming a family of topological space $(\alpha_t)_{t\in\iota}$, this distribution is a measure on the product-$\sigma$-field $\mathcal F :=\bigotimes_{t\in\iota} \mathcal B(\alpha_t)$ (where $\mathcal B(\alpha_t)$ is the Borel $\sigma$-algebra on $\alpha_t$). Here, $\mathcal F$ is generated by finite projections and hence any element of $\mathcal F$ may only depend on at most countably many $t\in \iota$, making this a rather coarse $\sigma$-algebra. (In particular, note that this is not the Borel $\sigma$-algebra of the product topology for infinite $\iota$.) The result responsible for this insight is usually denoted the {\em Kolmogorov extension theorem}, formulated in \cite{kolmogoroff1933grundbegriffe}. We note that a version of the extension result was proved by Daniell in the 1930s, but this paper was not acknowledged by the probabilists of that time \cite{aldrich2007but}. Due to his contribution, the theorem is often called the {\em Daniell-Kolmogorov extension}.

The goal of our contribution is a formalization of the proof of that theorem in {\tt mathlib4}, the mathematical library of {\tt Lean4} \cite{mathlib2020lean}. The Kolmogorov extension theorem is on the interface between measure theory and probability theory. Here, we rely on a decent amount of formalized mathematics in the measure-theory part of {\tt mathlib4} (outer measures, above all), while not requiring any specific previous formalization of probability theory. (In fact, most of our results are formulated in terms of finite rather than probability measures.)

@@

The main steps in our construction, which were previously missing in {\tt mathlib4}, are (more mathematical details are below, and in standard textbooks on Probability Theory, e.g.\ \cite{Klenke2013, Kallenberg2020}):
\begin{enumerate}
\item a formalization of (the set-system of) semi-rings (see Definition~\ref{def:semi} below); 
\item adding some main lemmas to measure theory, e.g.\ continuity of a measure at the emptyset (see e.g.\ Lemma~\ref{l:halbextRing}, Proposition~\ref{P:stetigmass}, Lemma~\ref{l:stetigcompact} below);
\item a proof that a single probability measure on a Polish space is inner regular with respect to compact sets (see Definition~\ref{def:innerreg}; in fact, we proved a slight generalization using pseudo-metric spaces, found in Lemma~\ref{L:relcoPol} below); 
\item the classical Carathéodory extension theorem, providing us a candidate for the measure which we want to construct (see Theorem~\ref{T:masseind}); 
\item the Kolmogorov extension theorem, as based on the previous steps (See Theorem~\ref{T1}).
\end{enumerate}
The Kolmogorov extension theorem has been previously formalized in Isabelle/HOL \cite{Immler2012}. This formalization only works on Polish spaces (rather than on spaces where every finite measure is inner regular with respect to compact sets, see below), and only in the case where all $\alpha_i$'s are identical. However, type theory as the basis of the formalization, as well as several mathematical concepts (e.g.\ projective families of measures or the continuity of a measure in the emptyset) are the same in both formalisations in these two interactive theorem provers.

\subsubsection*{Possible future work}
\sloppy Let us describe some future projects extending {\tt mathlib4} which become possible by our contribution.

\noindent
{\em Instances of stochastic processes:} An obvious application of Kolmogorov's extension theorem is the construction of basic stochastic processes like Markov chains \cite[Section~11]{Kallenberg2020}, the Poisson process \cite[Section~13]{Kallenberg2020} and Brownian motion \cite[Section~14]{Kallenberg2020}. (Note that other Gaussian processes -- indexed by $[0, \infty)$ or $\mathbb R$ -- might as well be constructed the same way. In addition, fields like the Gaussien free field indexed by $\mathbb R^d$ \cite{sheffield2007gaussian} can also be given.) For these tasks, we would have to define the finite-dimensional distributions (Poisson and normally distributed, respectively), and apply the extension theorem. This task requires the formalization of multi-dimensional Poisson and normal distributions, which -- in textbooks -- is usually done using characteristic functions. Since these are not yet part of {\tt mathlib4}, we postpone this task to the future.

\noindent
{\em Sample-path properties:} The Kolmogorov extension gives the existence of a distribution of a stochastic process $(X_t)_{t\in\iota}$ with certain properties. Extra work is needed in order to show that -- on the same probability space -- we can as well define a version (i.e.\ another process $(Y_t)_{t\in\iota}$ with $\mathbf P(X_t = Y_t) = 1$ for all $t\in\iota$) which is right-continuous with left limits (for the Poisson process) or continuous (for Brownian motion), respectively. For the former, this follows from some general principles of Markov processes (e.g.\ \cite[Theorem 4.3.6]{EthierKurtz1986}). For the latter, this requires formalization of the Kolmogorov-Chentsov criterion 
\cite[Theorem~4.23]{Kallenberg2020}.

\noindent
{\em Theorem of Ionescu-Tulcea and related results:} While Kolmogorov's extension Theorem gives a result for arbitrary $\iota$, but requires some properties of the family $(\alpha_t)_{t\in\iota}$, the Theorem of Ionescu-Tulcea \cite[Theorem 8.24]{Kallenberg2020} can only deal with countable $\iota$, but has no restrictions on $(\alpha_t)_{t\in\iota}$. The proof of the latter, however, uses transition kernels (and an inductive construction of a content, which can be extended to a measure) rather than projective families. However, continuity of a measure at $\emptyset$ is used as well and a formalization would complement our work. In addition, using the Theorem of Ionescu-Tulcea one can define infinitely many independent random variables (arbitrary $\iota$) on a joint probability space \cite[Theorem 8.24]{Kallenberg2020} without requirements on $(\alpha_t)_{t\in\iota}$. All of these results can be used in order to construct countably many independent random variables, which are frequently used in concrete constructions in probability theory. Examples are (one direction of) the Borel-Cantelli-Lemma \cite[Theorem 4.28]{Kallenberg2020}, Kolmogorov's 0-1-law \cite[Theorem 4.13]{Kallenberg2020}, random walks \cite[Chapter 12]{Kallenberg2020}, branching processes  \cite[Chapter 13]{Kallenberg2020}, and percolation  \cite[Chapter 2.4]{Klenke2013}, to name just a few. While the first two are already formalized in \leanline{mathlib4}, it is only clear due to the Kolmorov extension theorem (or the Theorem of Ionesu-Tulcea) that infinitely many independent random variables can indeed be defined on a joint probability space. In this sense, our contribution is also important for giving more sense to already formalised parts of \leanline{mathlib}.

\section{Formalization of the Daniell-Kolmogorov extension}
We start by stating the exact result. We are going to formulate the main result in a modern fashion, as e.g.\ found in Theorem 2.2 of \cite{rao1971projective}, Theorem 7.7.1 of Volume~2 of \cite{bogachev2007measure}, Theorem 15.26 of \cite{guide2006infinite}, or \cite{border1998expository}. Note that these formulations split general assumptions on the underlying space(s) (e.g.\ a metric property) from the property which is needed in the proof (inner regularity with respect to compact sets). Other -- highly readable -- references such as \cite{Billingsley1995} state the extension theorem only in special cases such as $\alpha_t = \mathbb R$ for all $t$. 
%In addition, there are still new proofs found for the general extension theorem \cite{chin2019new}, but they require some results not yet formalized in mathlib4 (Borel isomorphism of $\mathbb R$ to $\{0,1\}^{\mathbb N}$), so we do not follow this route.

\subsection{Formulating the result}
Before we can state the result, we begin with some mathematical notions. We start off from metric spaces, but quickly introduce projective families of measures. In all definitions, $\alpha$ will be some set (or type, since {\tt Lean4} is a dependently typed language, \leanline{α : Type _}).

\begin{definition}[Metric and topological spaces]
  \begin{enumerate}
  \item \sloppy A pseudo-metric on $\alpha$ is a symmetric map $r : \alpha
    \times \alpha \to [0,\infty)$ satisfying the triangle inequality,
      i.e.\ $r(x,z) \leq r(x,y) + r(y,z)$ for all $x,y,z\in\alpha$.
      \\ If $r$ also satisfies $r(x,y) = 0$ iff $x=y$, we call it a
      metric.  \\ If $r : \alpha \times \alpha \to [0,\infty]$
      (i.e.\ $r(x,y) = \infty$ is allowed), we call $r$ an extended
      (pseudo-)metric.
    \item Let $r$ be an (extended pseudo-)metric. A sequence $x_1,
      x_2,... \in \alpha$ is called Cauchy if for all $\varepsilon>0$
      there is $N\in\mathbb N$ such that $r(x_n, x_m) < \varepsilon$
      for all $m,n > N$. The (extended pseudo-) metric is called
      complete if every Cauchy-sequency has a limit in
      $\alpha$.
    \item Some\footnote{We denote by $2^\alpha$ the power set of
      $\alpha$, i.e.\ the set of all subsets of $\alpha$.} $\mathcal O
      \subseteq 2^\alpha$ is called a topology, if it satisfies (i)
      $\emptyset, \alpha \in \mathcal O$, (ii) $\mathcal O$ is a
      $\pi$-system, i.e.\ it is stable under finite intersections,
      i.e.\ if $A, B \in \mathcal O$, then $A \cap B \in \mathcal O$,
      and (iii) $\mathcal O$ is stable under arbitrary unions,
      i.e.\ if $A_i \in \mathcal O$ for all $i\in\iota$ and $\iota$ is
      arbitrary, then $\bigcup_{i\in \iota} A_i \in \mathcal O$.
  \end{enumerate}
\end{definition}

\begin{remark}[Metric and topological spaces in {\tt mathlib4}] \mbox{}
  \begin{enumerate}
    \item \sloppy In \leanline{Lean4}, we have the \leanline{class
      PseudoEMetricSpace (α : Type u) extends EDist}, where
      \leanline{EDist} comes with an extended distance \leanline{edist
        : α → α → ENNReal}, and the extension to
      \leanline{PseudoEMetricSpace} consists of the components,
      \leanline{edist_self}, \leanline{edist_comm},
      \leanline{edist_triangle} providing the properties of the
      pseudo-metric, and \leanline{toUniformSpace},
      \leanline{uniformity_edist} defining a uniform space from the
      extended pseudo-metric. Such a space does not come with a
      metric, but with a filter on $\alpha \times \alpha$, which
      describes which points in $\alpha$ are near. For example, the
      diagonal of $\alpha \times \alpha$ is a subset of all sets in
      the uniformity; see \cite{james2013topologies} for details. We
      note that a uniform space with a countably generated uniformity
      filter is pseudometrizable, i.e.\ there exists a
      pseudo-metric-space structure that generates the same
      uniformity; see \leanline{UniformSpace.pseudoMetrizableSpace},
      which formalizes a result stated in
      \cite{melikhov2011metrizable}.
    \item \sloppy For topological spaces, we have \leanline{class
      TopologicalSpace(α : Type u) : Type u}, which comes with
      \leanline{IsOpen : Prop}, \leanline{isOpen_univ : IsOpen
        Set.univ}, \leanline{isOpen_inter} and
      \leanline{isOpen_sUnion}, which are exactly the properties of a
      topological space described above.
    \item Any (extended pseudo-)metric on $\alpha$ defines a topology,
      namely the topology generated by $\mathcal H := \{\{y:
      r(x,y)<\varepsilon\}: x \in \alpha, \varepsilon \in
      (0,\infty)\}$. Actually, in \leanline{mathlib4}, as we have seen
      above, every extended pseudo-metric defines a uniform space, and
      a uniform space is a \leanline{class UniformSpace(α : Type u)
        extends TopologicalSpace , UniformSpace.Core} (where the
      uniformity is defined in \leanline{UniformSpace.Core}), which
      connects the \leanline{uniformity} with the topology using
      \leanline{∀ (s : Set α), IsOpen s ↔ ∀ (x : α), x ∈ s → {p |
          p.fst = x → p.snd ∈ s} ∈ uniformity}. In particular, a
      uniform space defines a topological space, which can be used by
      the typeclass system.
  \end{enumerate}
\end{remark}

\begin{remark}[Generated topology]
  \begin{enumerate}
  \item The intersection of any number of topologies is again a
    topology. For this reason, if $\mathcal H \subseteq 2^\alpha$, we
    define the topology $\mathcal O := \bigcap_{\mathcal F \supseteq
      \mathcal H \text{ topology}} \mathcal F$; see
    \leanline{TopologicalSpace.generateFrom}. This is called the
    topology generated by $\mathcal H$. If $\mathcal H$ is closed
    under finite intersections, we call $\mathcal H$ a basis for
    $\mathcal O$; see \leanline{TopologicalSpace.IsTopologicalBasis}.
  \item \sloppy We call the topology (generated from an extended
    pseudo-metric) separable (see
    \leanline{TopologicalSpace.SeparableSpace}) if there is a
    countable $s \subseteq \iota$ such that $\inf\{r(x,y) : y \in s\}
    = 0$ for all $x\in\alpha$. (More generally, a
    \leanline{TopologicalSpace α} is a \leanline{SeparableSpace} iff
    \leanline{∃ s, Set.Countable s ∧ Dense s}, where the latter means
    \leanline{∀ (x : α), x ∈ closure s}, i.e.\ the space equals its
    own closure.) \\ If there is a countable basis of the topology, it
    is separable; see
    \leanline{TopologicalSpace.SecondCountableTopology.to_separableSpace}.
  \end{enumerate}
\end{remark}

\noindent
Finally, we can introduce measures. A measure is defined on a
$\sigma$-algebra, which we introduce next.

\begin{definition}[$\sigma$-algebras and measures]\label{def:sigma}
  \begin{enumerate}
  \item 
    We call $\mathcal F \subseteq 2^\alpha$ a $\sigma$-algebra (on
    $\alpha$) if (i) $\emptyset \in \mathcal F$, (ii) $\mathcal F$ is
    stable under complements, i.e.\ $A \in \mathcal F \Rightarrow A^c
    \in \mathcal F$, (iii) $\mathcal F$ is stable under countable
    unions, i.e.\ $A_1, A_2,... \in \mathcal F \Rightarrow
    \bigcup_{n=1}^\infty A_n \in \mathcal F$. We call $(\alpha,
    \mathcal F)$ a measurable space.
  \item For some $\sigma$-algebra $\mathcal F$ on $\alpha$, a function
    $\mu : \mathcal F \to [0, \infty]$ is called a measure, if (i)
    $\mu(\emptyset) = 0$, (ii) $\mu$ is countably additive, i.e.\ for
    $A_1, A_2,... \in \mathcal F$ pairwise disjoint, we have $\mu\Big(
    \bigcup_{n=1}^\infty A_n\Big) = \sum_{n=1}^\infty \mu(A_n)$. We
    call $(\alpha, \mathcal F, \mu)$ a measure space. \\ In addition,
    $\mu$ is called finite if $\mu(\alpha) < \infty$ and a probability
    measure if $\mu(\alpha)=1$.
  \end{enumerate}
\end{definition}

\begin{remark}[Measur(abl)e spaces in \leanline{mathlib4}]
  \begin{enumerate}
  \item \sloppy The \leanline{class MeasurableSpace α} is very similar
    to a topological space in \leanline{mathlib4}, since it comes with
    \leanline{MeasurableSet' : Set α → Prop},
    \leanline{measurableSet_empty : MeasurableSet' ∅},
    \leanline{measurableSet_compl : (s : Set α) → MeasurableSet' s →
      MeasurableSet' sᶜ}, and \leanline{measurableSet_iUnion},
    i.e.\ properties (i)-(iii) from Definition~\ref{def:sigma}.1.
  \item We are using outer measures in our construction. In
    \leanline{mathlib4}, this is \leanline{structure
      MeasureTheory.OuterMeasure α} which is a function
    \leanline{measureOf : Set α → ℝ≥0∞} such that \leanline{measureOf
      ∅ = 0} (i.e.\ the empty set has measure~$0$), \leanline{∀ {s₁ s₂
        : Set α}, s₁ ⊆ s₂ → measureOf s₁ ≤ measureOf s₂}
    (i.e.\ monotonicity) and \leanline{∀ (s : ℕ → Set α), measureOf (⋃
      (i : ℕ), s i) ≤ ∑' (i : ℕ), measureOf (s i)} (which we call
    $\sigma$-subadditivity). \\ Note that an outer measure is defined
    on $2^\alpha$, whereas a measure is only defined on a subset (the
    $\sigma$-algebra of measurable sets). More precisely: A
    \leanline{MeasureSpace α} extends a \leanline{MeasurableSpace α}
    by a \leanline{Measure α} (which is usually called
    \leanline{volume}). For the measure, we have: \leanline{structure
      MeasureTheory.Measure α [MeasurableSpace α] extends
      MeasureTheory.OuterMeasure}. Here $\sigma$-additivity of a
    \leanline{Measure α} reads
\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  m_iUnion : ∀ ⦃f : ℕ → Set α⦄,
    (∀ (i : ℕ), MeasurableSet (f i)) →
    Pairwise (Disjoint on f) →
    ↑toOuterMeasure (⋃ (i : ℕ), f i) =
    ∑' (i : ℕ), ↑toOuterMeasure (f i)
\end{minted}    
  \end{enumerate}
\end{remark}

@@

\noindent
Using \leanline{variable {α : Type _} {C : Set (Set α)} {s t : Set α}
}, we define a semi-ring as follows:

@@

Let us remark that (i) \leanline{Lean4} indicates a coercion by
\leanline{↑}, which in this case is from \leanline{Finset} to
\leanline{Set} and (ii) we have \leanline{PairwiseDisjoint (s : Set ι)
  (f : ι → α)} iff the images of any distinct elements of \leanline{ι}
under \leanline{f} are different. (Hence, if \leanline{f = id}, the
usual definition of pairwise disjoint sets unfolds.)
